Stability and Chaos in Multi-agent Reinforcement Learning

### 2020-10-09 aamas2021-formatting-instructions/aamas21-formatting-instructions/sample.tex
minor mods
Updated introduction

\section{Introduction}

@@

An important result in modelling such multi-agent systems \fb{check
  consistency of multi agent $\to$ multi-agent.} from the EGT
perspective is that, when games are learnt and the assumptions of
rationality and perfect information are lifted, games may not converge
to an equilibrium. Instead, as shown in \cite{Sanders2018}, the
dynamics may be more complex, and even chaotic (see
Sec.~\ref{sec::DynamicalBehaviours}) \fb{Section $\to$ Sec.}. The present work further
describes that the emergence of such behaviours depends on the
parameters of the games and the learning algorithm. For the analysis
of multi-agent systems whose behaviour is inter-dependent, it is
desirable that the system be such that the game convergence to a
stable equilibrium. Without this, it is inherently impossible to
predict the outcome of learning.

\paragraph{Contribution}
In this study, we will be considering the question: under what choice
of parameters is a multi-agent system, which is trained on an iterated \fb{normal-form}
game using Q-Learning, likely to converge to an equilibrium as opposed
to displaying complex, or unstable behaviours? In investigating this
question, we will make the following assumptions:

@@

\fb{I would summarize here the main findings of the paper in 10-15 lines.}

@@

Analyses have been performed from the point of view of the replicator
dynamic towards the various types of behaviours that can emerge from
iterated games. Such games are considered imperfect, in that agents
make decisions by attempting to anticipate their opponents' behaviour
based on experience \cite{Galla2011}. \cite{Imhof2005} presents the
observation that, in the game of Prisoner's Dilemma, cyclic behaviour
emerges in which the optimal strategy cycles across defection,
cooperation and tit-for-tat despite defection being the
NE. \cite{Galla2011} builds on this observation by showing that, in fact,
such behaviour is highly dependent on the presence of memory loss in
the system. In the absence of memory loss, the learning converges to
the NE as expected. \fb{I think we can remove this last couple of lines if need be: However, as this parameter is increased, the
stability of the NE is removed and other stable points are created.}

Taking this notion \fb{which one?} even further, \cite{Galla2013}
describes an analysis of two-person iterated games in which the
players learn using \textit{experience weighted attraction} (EWA), a
form of reinforcement learning typically applied in experimental
economics \cite{Camerer2009}. Through numerical simulations, the
authors were able to show that the emergence of chaos and cycles is
dependent on the choice of parameters (see \cite{Strogatz2000} for
details on chaotic and periodic motion). A rigorous theoretical
analysis of the replicator dynamic corresponding to EWA, which applies
the techniques in \cite{Opper1992}, presented a method for
characterising the regions in parameter space in which a game is
likely to converge, exhibit chaos or limit cycles. More recently, the
same authors presented an extended study in \cite{Sanders2018}, in which
this analysis was extended towards a generic $p$-player game, which
showed that chaotic dynamics are more likely to be observed as the
number of players \textit{p} increases.

@@

The relation between RL and the replicator dynamic allows for the
assumption of convergence in reinforcement learning to be lifted and
instead for the emergence of more complex behaviour to be examined. In
particular, this study aims to perform a similar analysis to that in
\cite{Sanders2018} and establish the regions in parameter space in
which Q-Learning converges to stable fixed points and where the
learning is unstable. To the best of our knowledge, no such work
exists in this context.

\newpage

@@

    As discussed in the previous section, when learning on iterated
    games, player strategies may exhibit much more complex behaviours
    than convergence to a Nash Equilibrium (NE), including convergence
    to a unique equilibrium (though not always to an NE), convergence
    to one of multiple equilibria, limit cycles, and chaos. These
    behaviours are illustrated in
    Fig.\ref{fig::DynamicalBehaviours}. To be able to predict the
    behaviour of a learning algorithm, it should ideally converge to a
    stable equilibrium although it is still possible to study systems
    with multiple equilibria or limit cycles
    \cite{Strogatz2000}. However, it would be difficult to control
    systems whose dynamics are governed by chaos (though research into
    controlling chaos is ongoing and rife with opportunity
    \cite{Fradkov2009}). It would, therefore, be a useful endeavour to
    determine the conditions under which the different sorts of behaviours
    arise.

@@

We now scale the system so that the probabilities sum to $N$. The
reason for this scaling is that, as the number of agents $p$ and
number of actions $N$ increases, the expected payoffs that agent
$\xmu{i}{\mu}$ receives all converge to the same value across its action
space. Therefore, the system must be scaled to ensure appreciable
differences across actions, so that the agent can choose its actions
appropriately (see the discussion in \cite{Sanders2018} or
Section S2.2 for further details). Accordingly we choose
scaled variables (denoted with a tilde) as
%
\begin{eqnarray*}
%    \begin{split}
        \payoff{i}{\mu} & = & \tpayoff{i}{\mu} \sqrt{N^{p-1}}\\
        \xmu{i}{\mu} & = & \frac{\txmu{i}{\mu}}{N} \\
        \talpha & = & \frac{\alpha}{N} \\
        \ttau & = & N^{-(p-1)/2} \tau \\
        \htau & = & N^{-p/2} \tau.
%    \end{split}
\end{eqnarray*}

@@

As mentioned, we now
%wish to
generalise these dynamics to account for all the possible realisations
of the payoff matrix. To do this, we assert that the payoff elements
(after scaling) are generated by a multivariate Gaussian distribution
with mean zero and covariance given as
%
\begin{equation}
\label{eqn::Payoffs}
%    \begin{split}
        \mathbb{E}\left [ \payoff{i}{\mu} \payoff{i}{\nu} \right] = \begin{cases}
        \frac{1}{N^{p-1}} &  \text{ if } \nu = \mu \\
        \frac{\Gamma}{(p-1) N^{p-1}} & \text{ otherwise. }
        \end{cases}
%    \end{split}
\end{equation}

@@

We then use a generating functional approach as outlined in
\cite{Mezard1986} to average the dynamics over this multivariate
Gaussian. The generating functional is a method derived from
statistical mechanics which allows for expectations to be taken over
equations of motion and, in particular, allows for the analysis of
systems with `quenched disorder': these are
variables which are random but held fixed in time as the system
evolves (in our case these variables are the payoff elements). The
generating functional is given as \cite{Mezard1986}

@@

where the equations of motion are the Lagrange equations of motion
given in ( \ref{eqn::scaledEOM}) and the fields $\Vec{\psi_i(t)}$ and
$\Vec{\phi_i(t)}$ will be set to zero at the end of the
calculation. $\delta$ denotes the Dirac delta function. Substituting
for our equation of motion yields
%
{\small
\begin{equation}
\label{eqn::separatePayoffs}
	\begin{split}
	Z(\Vec{\psi}) = \int D[\Vec{x}, \Vec{\hat{x}}] exp( i \sum_{i, \mu} \int dt [ \hxmu{i} (\frac{\dxmu{i}}{\xmu{i}{\mu}} - \talpha \rho_i^\mu (t) - h_i^\mu (t))]) \times \\ exp(-i \alpha \ttau \sum_{\mu} \sum_{i_\mu, i_{-\mu}} \int dt [\hxmu{i} \payoff{i}{\mu} \prod_{\kappa \neq \mu} \xmu{i}{\kappa} )])) 
    \times \\ exp(-i \alpha \ttau \sum_{\mu} \sum_{j_\mu, i_\mu, i_{-\mu}} \int dt [\hxmu{j}  \xmu{i}{\mu} \payoff{i}{\mu} \prod_{\kappa \neq \mu} \xmu{i}{\kappa}])) 
	\times \\ exp(i \sum_{i, \mu}
	\int dt[\xmu{i} \psi^\mu_i(t)]),
\end{split}
\end{equation}
}
The last two exponentials contain the payoffs of the game. These are randomly generated using a multivariate gaussian and then held fixed for the rest of the game. As such these comprise the aforementioned 'quenched disorder'. We will average over this quenched disorder (and therefore average over all possible realisations of payoff matrices) using the mean and covariance expressions given in (\ref{eqn::Payoffs}). We define $Q$ to be the product of the second and third exponentials in (\ref{eqn::separatePayoffs}) 
%

@@

By taking this average (Section S2.3), we obtain the \textit{effective dynamics}

@@

We then examine the
long-term behaviour of the perturbation $\xpert$. by taking
the Fourier transform of (\ref{eqn::Linearised}) and analysing its
behaviour at $\omega = 0$. After some manipulation (Section S4), this gives

@@

The fixed point condition (\ref{eqn::Final}) yields a number of testable implications, of which we test the validity through in the subsequent section. These implications are as follows.

@@

We consider each of the implications (2)-(5) from the previous section
in turn, choosing to forego illustrating (1) as it can be seen immediately
from (\ref{eqn::Final}). $\tau$ is held fixed
at $0.05$ and vary $\Gamma$ in
the range $[-1, 0]$, $\alpha$ in the range $[0.01, 0.05]$.

@@

In this section, we describe and discuss the numerical experiments we use to verify the implications described in the discussion of Section \ref{sec::Theory}. We first describe how these experiments were conducted and then, having presented the results in Figures \ref{fig:NumericalExperiments} and \ref{fig:pVarianceExpt}, we discuss the correlation with the theory.

@@

\begin{figure}[t]
    \includegraphics[width = 1.1 \linewidth, center]{Figures/a5e2var.png}
    \caption{The trajectories of action probabilities for 2 players with 5 actions trained on a game of $\Gamma = -0.5$ and $\tau = 0.05$. (Left) $\alpha = 0.01$, (Right) $\alpha = 0.05$.}
    \label{fig:AlphaVariation}
\end{figure}

>>>>

\section{Introduction}
\label{sec::Intro}

@@

An important result in modelling such multi-agent systems from the EGT
perspective is that, when games are learnt and the assumptions of
rationality and perfect information are lifted, games may not converge
to an equilibrium. Instead, as shown in \cite{Sanders2018}, the
dynamics may be more complex, and even chaotic (see
Sec.~\ref{sec::DynamicalBehaviours}). The present work further
describes that the emergence of such behaviours depends on the
parameters of the games and the learning algorithm. For the analysis
of multi-agent systems whose behaviour is inter-dependent, it is
desirable that the system be such that the game convergence to a
stable equilibrium. Without this, it is inherently impossible to
predict the outcome of learning.

\paragraph{Contribution}
In this study, we will be considering the question: under what choice
of parameters is a multi-agent system, which is trained on an iterated normal-form game using Q-Learning, likely to converge to an equilibrium as opposed
to displaying complex, or unstable behaviours? In investigating this
question, we will make the following assumptions:

@@

Through our analysis and experiments we find that likelihood of displaying unstable dynamics increases as the \textit{step length} $\alpha$ and \textit{intensity of choice} $\tau$ parameters in the Q-Learning algorithm \cite{Barber2012} increase. However, the correlation between payoffs $\Gamma$, which measures how cooperative or competitive a game is, does not affect the stability of the system. In addition, we find that regardless of the choice of these parameters, the likelihood of unstable behaviours increases as the number of players $p$ in the game increase.

@@

Analyses have been performed from the point of view of the replicator
dynamic towards the various types of behaviours that can emerge from
iterated games. Such games are considered imperfect, in that agents
make decisions by attempting to anticipate their opponents' behaviour
based on experience \cite{Galla2011}. \cite{Imhof2005} presents the
observation that, in the game of Prisoner's Dilemma, cyclic behaviour
emerges in which the optimal strategy cycles across defection,
cooperation and tit-for-tat despite defection being the
NE. In \cite{Galla2011}, Galla builds on this observation by showing that, in fact,
such behaviour is highly dependent on the presence of memory loss in
the system. In the absence of memory loss, the learning converges to
the NE as expected.However, as this parameter is increased, the
stability of the NE is removed and other stable points are created.

Taking the notion of parameter dependence even further, \cite{Galla2013}
describes an analysis of two-person iterated games in which the
players learn using \textit{experience weighted attraction} (EWA), a
form of reinforcement learning typically applied in experimental
economics \cite{Camerer2009}. Through numerical simulations, the
authors were able to show that the emergence of chaos and cycles is
dependent on the choice of parameters (see \cite{Strogatz2000} for
details on chaotic and periodic motion). A rigorous theoretical
analysis of the replicator dynamic corresponding to EWA, which applies
the techniques in \cite{Opper1992}, presented a method for
characterising the regions in parameter space in which a game is
likely to converge, exhibit chaos or limit cycles. More recently, the
same authors presented an extended study in \cite{Sanders2018}, in which
this analysis was extended towards a generic $p$-player game, which
showed that chaotic dynamics are more likely to be observed as the
number of players \textit{p} increases. By contrast to the present work, \cite{Barber2012} focuses on the analysis of the EWA algorithm, while we consider the Q-Learning \cite{Barber2012} algorithm.

@@

The relation between RL and the replicator dynamic allows for the
assumption of convergence in reinforcement learning to be lifted and
instead for the emergence of more complex behaviour to be examined. In
particular, this study aims to perform a similar analysis to that in
\cite{Sanders2018} and establish the regions in parameter space in
which Q-Learning converges to stable fixed points and where the
learning is unstable. In doing so, we extend the dynamics considered by Tuyls et al. towards a general $p$-player scenario, rather than only two players, thereby allowing an analysis of how the number of players in a game affects its stability. To the best of our knowledge, no such work
exists in this context.

@@

    As discussed in the Sec. \ref{sec::Intro}, when learning on iterated
    games, player strategies may exhibit much more complex behaviours
    than convergence to a Nash Equilibrium (NE), including convergence
    to a unique equilibrium (though not always to an NE), convergence
    to one of multiple equilibria, limit cycles, and chaos. These
    behaviours are illustrated in
    Fig.\ref{fig::DynamicalBehaviours}. To be able to predict the
    behaviour of a learning algorithm, it should ideally converge to a
    stable equilibrium although it is still possible to study systems
    with multiple equilibria or limit cycles
    \cite{Strogatz2000}. However, it would be difficult to control
    systems whose dynamics are governed by chaos (though research into
    controlling chaos is ongoing and rife with opportunity
    \cite{Fradkov2009}). It would, therefore, be a useful endeavour to
    determine the conditions under which the different sorts of behaviours
    arise.

@@

We now scale the system so that the probabilities sum to $N$. The
reason for this scaling is that, as the number of agents $p$ and
number of actions $N$ increases, the expected payoffs that agent
$\xmu{i}{\mu}$ receives all converge to the same value across its action
space. Therefore, the system must be scaled to ensure appreciable
differences across actions, so that the agent can choose its actions
appropriately (see the discussion in \cite{Sanders2018} or
Sec. S2.2 for further details). Accordingly we choose
scaled variables (denoted with a tilde) as
%
\begin{eqnarray*}
%    \begin{split}
        \payoff{i}{\mu} & = & \tpayoff{i}{\mu} \sqrt{N^{p-1}}\\
        \xmu{i}{\mu} & = & \frac{\txmu{i}{\mu}}{N} \\
        \talpha & = & \frac{\alpha}{N} \\
        \ttau & = & N^{-(p-1)/2} \tau \\
        \htau & = & N^{-p/2} \tau.
%    \end{split}
\end{eqnarray*}

@@

As mentioned, we now
%wish to
generalise these dynamics to account for all the possible realisations
of the payoff matrix. To do this, we assert that the payoff elements
(after scaling) are generated by a multi-variate Gaussian distribution
with mean zero and covariance given as
%
\begin{equation}
\label{eqn::Payoffs}
%    \begin{split}
        \mathbb{E}\left [ \payoff{i}{\mu} \payoff{i}{\nu} \right] = \begin{cases}
        \frac{1}{N^{p-1}} &  \text{ if } \nu = \mu \\
        \frac{\Gamma}{(p-1) N^{p-1}} & \text{ otherwise. }
        \end{cases}
%    \end{split}
\end{equation}

@@

We then use a generating functional approach as outlined in
\cite{Mezard1986} to average the dynamics over this multi-variate
Gaussian. The generating functional is a method derived from
statistical mechanics which allows for expectations to be taken over
equations of motion and, in particular, allows for the analysis of
systems with `quenched disorder': these are
variables which are random but held fixed in time as the system
evolves (in our case these variables are the payoff elements). The
generating functional is given as \cite{Mezard1986}

@@

where the equations of motion are the Lagrange equations of motion
given in ( \ref{eqn::scaledEOM}) and the fields $\Vec{\psi_i(t)}$ and
$\Vec{\phi_i(t)}$ will be set to zero at the end of the
calculation. $\delta$ denotes the Dirac delta function. Substituting
for our equation of motion yields
%
{\small
\begin{equation}
\label{eqn::separatePayoffs}
	\begin{split}
	Z(\Vec{\psi}) = \int D[\Vec{x}, \Vec{\hat{x}}] exp( i \sum_{i, \mu} \int dt [ \hxmu{i} (\frac{\dxmu{i}}{\xmu{i}{\mu}} - \talpha \rho_i^\mu (t) - h_i^\mu (t))]) \times \\ exp(-i \alpha \ttau \sum_{\mu} \sum_{i_\mu, i_{-\mu}} \int dt [\hxmu{i} \payoff{i}{\mu} \prod_{\kappa \neq \mu} \xmu{i}{\kappa} )])) 
    \times \\ exp(-i \alpha \ttau \sum_{\mu} \sum_{j_\mu, i_\mu, i_{-\mu}} \int dt [\hxmu{j}  \xmu{i}{\mu} \payoff{i}{\mu} \prod_{\kappa \neq \mu} \xmu{i}{\kappa}])) 
	\times \\ exp(i \sum_{i, \mu}
	\int dt[\xmu{i} \psi^\mu_i(t)]),
\end{split}
\end{equation}
}
The last two exponentials contain the payoffs of the game. These are randomly generated using a multi-variate gaussian and then held fixed for the rest of the game. As such these comprise the aforementioned 'quenched disorder'. We will average over this quenched disorder (and therefore average over all possible realisations of payoff matrices) using the mean and covariance expressions given in (\ref{eqn::Payoffs}). We define $Q$ to be the product of the second and third exponentials in (\ref{eqn::separatePayoffs}) 
%

@@

By taking this average (Sec. S2.3), we obtain the \textit{effective dynamics}

@@

We then examine the
long-term behaviour of the perturbation $\xpert$. by taking
the Fourier transform of (\ref{eqn::Linearised}) and analysing its
behaviour at $\omega = 0$. After some manipulation (Sec. S4), this gives

@@

The fixed point condition (\ref{eqn::Final}) yields a number of testable implications, of which we subsequently test the validity. These implications are as follows.

@@

We consider each of the implications (2)-(5) from the Sec. \ref{sec::Theory}
in turn, choosing to forego illustrating (1) as it can be seen immediately
from (\ref{eqn::Final}). $\tau$ is held fixed
at $0.05$ and vary $\Gamma$ in
the range $[-1, 0]$, $\alpha$ in the range $[0.01, 0.05]$.

@@

In this section, we describe and discuss the numerical experiments we use to verify the implications described in the discussion of Sec. \ref{sec::Theory}. We first describe how these experiments were conducted and then, having presented the results in Figures \ref{fig:NumericalExperiments} and \ref{fig:pVarianceExpt}, we discuss the correlation with the theory.

@@

\begin{figure}[t]
    \includegraphics[width = 0.75 \linewidth, center]{Figures/alphavar5e2.png}
    \caption{The trajectories of action probabilities for 2 players with 5 actions trained on a game of $\Gamma = -0.5$ and $\tau = 0.05$. (Left) $\alpha = 0.01$, (Right) $\alpha = 0.05$.}
    \label{fig:AlphaVariation}
\end{figure}
